---
title: "Lesson 2: Objects in R"
output: 
  html_document:
    toc: FALSE
---

<style type="text/css">

body, td {
   font-size: 18px;
}
code.r{
  font-size: 14px;
}
pre {
  font-size: 14px
}
</style>

*This is the course handbook for WolfWorks: An introduction to R.*

***

Objectives:

1. Understand the key terms *object*, *variable*, *assign*, *call*, *function*, *argument*, and *options*
2. Know how to store and name objects within the RStudio Environment


***


## Assigning values to objects/variables

As we saw previously, we can execute code directly via the **console** 
(by pressing `Enter`) or indirectly via a **script** (by pressing `Control + Enter`).
The output of our command will appear in the console. However, if we want to 
store a value or data structure, we have to *assign* it to an *object*.
<br>
<br>

Some more important definitions: 

* An *object* is a data store within R - it contains one or more values and these can be stored in various structures e.g., data frames, matrices, vectors (more about these data structures later)
* What are known as `objects` in `R` are known as `variables` in some other programming languages. In certain scenarios the terms *object* and *variable* have very distinct meanings, but for the context of today's workshop they are the same thing
* To store some data within an object, we must *assign* the data to our object. This is done using the assignment operator `<-`
<br>

For example, if I wish to save the value `55` into an object called `weight_kg`,
I could use the assignment operator like so:


```{r}
# assign a value of 55 to a variable called weight_kg
weight_kg <- 55
```


When I execute this code I notice that there is no output in the console. That 
is because R does not print anything when assigning a value to an object. 
However, we do see that this object has appear in our Environment pane. If we
want R to both assign the value of `55` to `weight_kg` and at the same time print
its value in the console, we can add brackets around our assignment. 


```{r}
(weight_kg <- 55)
```


Now R has stored `weight` in its memory, we can use the object name to 
represent the value we have stored.


```{r}
# print the value of weight
weight_kg

# use the value of weight for arithmetic
weight_kg * 2.2
```


As before, the output that appears in our console is the result of the arithmetic
alongside a `[1]` to indicate that this is the first (and only) value of the output.
Importantly, this has not altered the value of our `weight_kg` object because we
did not assign the output of this code to the object `weight_kg`. If we want to 
save the value of `weight_kg * 2.2` then we need to assign it to an object.


```{r}
## assign the output of our arithmetic query to a new object
weight_lb <- weight_kg * 2.2
```


Now I see both objects are present in my environment. 

If we were to assign a new value to `weight_kg`, this would overwrite the 
previous value. Let's try it. 


```{r}
weight_kg <- 62
```


**Question:**
What are the values of `weight_kg` and `weight_lb` now. Does `weight_lg` = 
`weight_kg * 2.2`? 
<br>
<br>

**Naming objects in R**

Since we now know how to assign a value to an object using the `<-` operator, it
is worth taking a moment to consider how we should name these objects. There are
some basic naming principles that you should adhere to in R:

1. Use meaningful names that relate to what is inside of the object - avoid meaningless names such as `a`, `b`, `c`
2. R will not allow names with spaces in - it is common practice to replace a space with `_` 
3. R will not allow an object name which begins with a number - `2x` is not a valid name, but `x2` is
4. Avoid names that are the same as function names e.g., `if`, `else`, `mean`, `data` and `c` 
5. Avoid dots (`.`) as many function names contain dots for historical reasons

You should also be aware that R is case-sensitive. This means that R does not 
consider `weight_kg` and `Weight_kg` to be the same thing. For more information
about naming practices and writing neat code there are several R style guides 
e.g., the [Bioconductor style guide](https://contributions.bioconductor.org/r-code.html#r-code)
or the [tidyverse guide](https://style.tidyverse.org).
<br>
<br>

**Comments**

When we are writing a script we want to be able to annotate the script with 
notes and explanations of what the code is doing. This helps both your future 
self and anyone else who should ready your script to understand what is going
on.

The comment character in R is `#`. Anything to the right of a `#` will be 
ignored by R. RStudio also helps us by changing the colour of our commented text
so that we can see it easily.


***


## Calling functions 

Functions are one of the key features of R. A *function* is a self-contained
module of code that has been written to carry out a specific task. R has many 
functions that allow us to automate common tasks. For example, think about how
many R users around the globe will at some point take the mean average of some 
data. Rather than each individual writing out the arithmetic for this, R has a
convenient `mean` function that already contains all of the required code.

Many functions are pre-defined in R and are already here and ready for use. For 
more specific analysis needs, thousands more functions can be installed and used
by importing R *packages* (more on that later).

A *function* typically requires one or more inputs called *arguments* and usually
returns a value. Executing a function (i.e., running it) is referred to as 
*calling* the function.

Let's look at a simple example, the `round` function. This function takes a
number and rounds it, as indicated by the name.


```{r}
round(x = 3.1415926)
```


Here, we have *called* the `round` *function* and passed it the argument 
`x = 3.1415926`. If we want to see what the argument `x` requires, we can use 
the single question mark help function that we saw previously.


```{r}
?round
```


This tells us that the argument `x` is a numeric vector i.e., the number that 
we wish to round. We can also see that there is another argument available for
this function, the `digits` argument. We did not previously pass this argument
because it is not an absolute requirement. Many functions have these optional
arguments (called *options*) and if they are not specified they will take on a 
*default* value. Here, the default value for the `digits` argument is 0 (i.e.,
round to the nearest whole number). The default value can be overwritten by 
specifying this argument in our code.


```{r}
round(x = 3.1415926, digits = 2)
```


In this example we have explicitly named the arguments `x =` and `digits =`. 
This is not always necessary in R, but it is useful when starting out. When we
name the arguments, the order we provide them in does not matter because R can
still tell what we are referring to. If, however, we provided our arguments 
without naming them, then we would have to be careful about their order.

There is a default order in which R expects to receive arguments for a function.
If we don't provide explicit argument names, we have to stick to the default 
order so that R can tell which argument is which.


```{r}
# Pass arguments in the correct (default) order with names
round(x = 3.1415926, digits = 2)

# Pass arguments in the correct (default) order without names
round(3.1415926, 2)

# Pass arguments in alternative order with names
round(digits = 2, x = 3.1415926)

# Pass arguments in alternative order without names
round(2, 3.1415926)
```


This is particularly important as we begin to use more complex functions that 
require a larger number of arguments.


Some useful math/stat functions in R: 

* `max()`: maximum value in a numeric vector
* `min()`: minimum value in a numeric vector
* `range()`: vector of min and max
* `sum()`: sum of a vector
* `mean()`: mean of a vector
* `median()`: median of a vector
* `var()`: variance of a vector
* `sd()`: standard deviation of a vector
* `sort()`: sorted version of a vector
* `length()`: length of an object
* `cor()`: correlation of x and y
<br>
<br>

**Challenge: Objects in R**
<br>
Create two new objects called `mass` and `age` and assign the values of `122`
and `47.5` to them, respectively. Now use these objects to calculate the value 
of a new object called `mass_index` (equal to mass divided by age).

Now change the value of `mass` by multiplying it by two and change the value of
`age` by minusing 20. What is the value of the `mass_index` now?

<details>  
<summary>Solution</summary> 
<br>

```{r}
mass <- 122
age <- 47.5

mass_index <- 122 / 47.5

mass <- mass * 2
age <- age - 20

mass_index
```

</details>